Two subtle issues with over-representation analysis: Demonstration

### 2024-07-01 app/report.Rmd
working

suppressPackageStartupMessages({
  library("clusterProfiler")
  library("markdown")
  library("rmarkdown")
  library("shiny")
  library("broom")
  library("RhpcBLASctl")
  library("eulerr")
  library("plotly")
  library("DT")
})

@@

ora$ES <- gr/br
ora$ID = ora$geneID = ora$p.adjust = ora$Count = NULL
colnames(ora) <- gsub("qvalue","FDR",colnames(ora))
colnames(ora) <- gsub("GeneRatio","FgRatio",colnames(ora))
ora <- ora[,c("Description","FgRatio","BgRatio","ES","pvalue","FDR")]
ora$ES <- signif(ora$ES,3)
ora$pvalue <- signif(ora$pvalue,3)
ora$FDR <- signif(ora$FDR,3)
ora1 <- ora

@@

DT::datatable(
  { ora1 },

@@

@@

  ora$ES <- gr/br
  ora$ID = ora$geneID = ora$p.adjust = ora$Count = NULL
  colnames(ora) <- gsub("qvalue","FDR",colnames(ora))
  colnames(ora) <- gsub("GeneRatio","FgRatio",colnames(ora))
  ora <- ora[,c("Description","FgRatio","BgRatio","ES","pvalue","FDR")]
  ora$ES <- signif(ora$ES,3)
  ora$pvalue <- signif(ora$pvalue,3)
  ora$FDR <- signif(ora$FDR,3)
} else if (comparison == "FDR error") {
  gs1 <- mygs
  gs1 <- gs1[which(gs1$gene %in% bg),]
  terms <- names(which(table(gs1$term)>5))
  gs1 <- gs1[gs1$term %in% terms,]
  options(enrichment_force_universe = FALSE)
  ora <- as.data.frame(enricher(gene = fg ,
                                universe = bg, minGSSize = 5, maxGSSize = 500000, TERM2GENE = gs1,
                                pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

@@

  ora$ES <- gr/br
  ora$ID = ora$geneID = ora$p.adjust = ora$Count = NULL
  colnames(ora) <- gsub("qvalue","FDR",colnames(ora))
  colnames(ora) <- gsub("GeneRatio","FgRatio",colnames(ora))
  ora <- ora[,c("Description","FgRatio","BgRatio","ES","pvalue","FDR")]

  # proper FDR correction
  nsets <- length(which(table(gs1$term)>5))
  nres <- nrow(ora)
  diff <- nsets - nres
  pvals <- c(ora$pvalue,rep(1,diff))
  ora$FDR <- p.adjust(pvals,method="fdr")[1:nrow(ora)]
}

ora2 <- ora
```

@@

DT::datatable(
  { ora2 },

@@

@@

```{r,comparison2}

@@

if ( comparison == "Background error") {
  v1 <- list("Original"=ora1_sets, "BG fix"=ora2_sets)
} else if (comparison == "FDR error") {
  v1 <- list("Original"=ora1_sets, "FDR fix"=ora2_sets)
}
plot(euler(v1),quantities = list(cex = 2), labels = list(cex = 2))

@@

  plot_ly(
    m, x = ~ES.x, y = ~ES.y
  ) %>%
    add_trace(m, x = ~Original, y = ~FDRcorrected, type = "scatter",
              mode="lines", line=list(color='grey')) %>%
    add_trace(m, x = ~ES.x, y = ~ES.y, type = "scatter", mode = "markers",
              showlegend = FALSE, text = m$Description,
              hoverinfo = 'text') %>%
    layout(title="Enrichment score comparison")
}

@@

  plot_ly(
    m, x = ~logFDR.x, y = ~logFDR.y
  ) %>%
    add_trace(m, x = ~Original, y = ~FDRcorrected, type = "scatter",
              mode="lines", line=list(color='grey')) %>%
    add_trace(m, x = ~logFDR.x, y = ~logFDR.y, type = "scatter", mode = "markers",
              showlegend = FALSE, text = m$Description,
              hoverinfo = 'text') %>%
    layout(title="log FDR comparison")
}

>>>>

suppressPackageStartupMessages({
  library("clusterProfiler")
  library("markdown")
  library("rmarkdown")
  library("shiny")
  library("broom")
  library("RhpcBLASctl")
  library("eulerr")
  library("plotly")
  library("DT")
  library("kableExtra")
})

@@

ora$ES <- gr/br
ora1_full <- ora
ora$ID = ora$geneID = ora$p.adjust = ora$Count = NULL
colnames(ora) <- gsub("qvalue","FDR",colnames(ora))
colnames(ora) <- gsub("GeneRatio","FgRatio",colnames(ora))
ora <- ora[,c("Description","FgRatio","BgRatio","ES","pvalue","FDR")]
ora$ES <- signif(ora$ES,3)
ora$pvalue <- signif(ora$pvalue,3)
ora$FDR <- signif(ora$FDR,3)
ora1 <- ora

@@

DT::datatable(
  { head(ora1,1000) },

@@

```{r,defaultfull,results="asis",warning=FALSE}

HEADER=paste("<br><details><summary><b>","Click HERE to show results for top 1000 sets","</b></summary><br><p>",sep=" " )

cat(HEADER)

ora1_full$ID = ora1_full$p.adjust = NULL
colnames(ora1_full) <- gsub("qvalue","FDR",colnames(ora1_full))
ora1_full$geneID <- gsub("/"," ",ora1_full$geneID)
ora1_full <- ora1_full[,c("Description","GeneRatio","BgRatio","Count","ES","pvalue","FDR","geneID")]
kbl(head(ora1_full,1000), format="html", row.names=FALSE, caption = "Complete results",digits=100) %>%
kable_styling("hover", full_width = FALSE)

cat("<br></p></details>")
cat("<hr><br>")
```

@@

  ora$ES <- gr/br
  ora2_full <- ora
  ora$ID = ora$geneID = ora$p.adjust = ora$Count = NULL
  colnames(ora) <- gsub("qvalue","FDR",colnames(ora))
  colnames(ora) <- gsub("GeneRatio","FgRatio",colnames(ora))
  ora <- ora[,c("Description","FgRatio","BgRatio","ES","pvalue","FDR")]
  ora$ES <- signif(ora$ES,3)
  ora$pvalue <- signif(ora$pvalue,3)
  ora$FDR <- signif(ora$FDR,3)
} else if (comparison == "FDR error") {
  gs1 <- mygs
  gs1 <- gs1[which(gs1$gene %in% bg),]
  terms <- names(which(table(gs1$term)>5))
  gs1 <- gs1[gs1$term %in% terms,]
  options(enrichment_force_universe = FALSE)
  ora <- as.data.frame(enricher(gene = fg ,
                                universe = bg, minGSSize = 5, maxGSSize = 500000, TERM2GENE = gs1,
                                pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

@@

  ora$ES <- gr/br
  ora2_full <- ora
  ora$ID = ora$geneID = ora$p.adjust = ora$Count = NULL
  colnames(ora) <- gsub("qvalue","FDR",colnames(ora))
  colnames(ora) <- gsub("GeneRatio","FgRatio",colnames(ora))
  ora <- ora[,c("Description","FgRatio","BgRatio","ES","pvalue","FDR")]

  # proper FDR correction
  nsets <- length(which(table(gs1$term)>5))
  nres <- nrow(ora)
  diff <- nsets - nres
  pvals <- c(ora$pvalue,rep(1,diff))
  ora$FDR <- p.adjust(pvals,method="fdr")[1:nrow(ora)]
  ora$ES <- signif(ora$ES,3)
  ora$pvalue <- signif(ora$pvalue,3)
  ora$FDR <- signif(ora$FDR,3)
} else if (comparison == "Both errors") {
  gs1 <- mygs
  gs1 <- gs1[which(gs1$gene %in% bg),]
  terms <- names(which(table(gs1$term)>5))
  gs1 <- gs1[gs1$term %in% terms,]
  options(enrichment_force_universe = TRUE)
  ora <- as.data.frame(enricher(gene = fg ,
                                universe = bg, minGSSize = 5, maxGSSize = 500000, TERM2GENE = gs1,
                                pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

  ora$ES <- gr/br
  ora2_full <- ora
  ora$ID = ora$geneID = ora$p.adjust = ora$Count = NULL
  colnames(ora) <- gsub("qvalue","FDR",colnames(ora))
  colnames(ora) <- gsub("GeneRatio","FgRatio",colnames(ora))
  ora <- ora[,c("Description","FgRatio","BgRatio","ES","pvalue","FDR")]

  # proper FDR correction
  nsets <- length(which(table(gs1$term)>5))
  nres <- nrow(ora)
  diff <- nsets - nres
  pvals <- c(ora$pvalue,rep(1,diff))
  ora$FDR <- p.adjust(pvals,method="fdr")[1:nrow(ora)]
  ora$ES <- signif(ora$ES,3)
  ora$pvalue <- signif(ora$pvalue,3)
  ora$FDR <- signif(ora$FDR,3)
}
ora2 <- ora
```

@@

DT::datatable(
  { head(ora2,1000) },

@@

```{r,fixedfull,results="asis"}

HEADER=paste("<br><details><summary><b>","Click HERE to show results for top 1000 sets","</b></summary><br><p>",sep=" " )

cat(HEADER)

ora2_full$ID = ora2_full$p.adjust = NULL
colnames(ora2_full) <- gsub("qvalue","FDR",colnames(ora2_full))
ora2_full$geneID <- gsub("/"," ",ora2_full$geneID)
ora2_full <- ora2_full[,c("Description","GeneRatio","BgRatio","Count","ES","pvalue","FDR","geneID")]

kbl(head(ora2_full,1000), format="html", row.names=FALSE, caption = "Complete results",digits=100) %>%
kable_styling("hover", full_width = FALSE)

cat("<br></p></details>")
cat("<hr><br>")

@@

```{r,comparison2,fig.height=5}

@@

if ( comparison == "Background error") {
  v1 <- list("Original"=ora1_sets, "BG fix"=ora2_sets)
} else if (comparison == "FDR error") {
  v1 <- list("Original"=ora1_sets, "FDR fix"=ora2_sets)
} else if (comparison == "Both errors") {
  v1 <- list("Original"=ora1_sets, "Both fix"=ora2_sets)
}
plot(euler(v1),quantities = list(cex = 2), labels = list(cex = 2))

@@

  plot_ly(
    m, x = ~ES.x, y = ~ES.y
  ) %>%
    add_trace(m, x = ~Original, y = ~FDRcorrected, type = "scatter",
              mode="lines", line=list(color='grey')) %>%
    add_trace(m, x = ~ES.x, y = ~ES.y, type = "scatter", mode = "markers",
              showlegend = FALSE, text = m$Description,
              hoverinfo = 'text') %>%
    layout(title="Enrichment score comparison")
} else if (comparison == "Both errors") {
  m$Original <- m$BOTHcorrected <- round(VEC[1:nrow(m)],0.1)
  m$ES.x <- signif(m$ES.x,3)
  m$ES.y <- signif(m$ES.y,3)

  plot_ly(
    m, x = ~ES.x, y = ~ES.y
  ) %>%
    add_trace(m, x = ~Original, y = ~BOTHcorrected, type = "scatter",
              mode="lines", line=list(color='grey')) %>%
    add_trace(m, x = ~ES.x, y = ~ES.y, type = "scatter", mode = "markers",
              showlegend = FALSE, text = m$Description,
              hoverinfo = 'text') %>%
    layout(title="Enrichment score comparison")
}

@@

  plot_ly(
    m, x = ~logFDR.x, y = ~logFDR.y
  ) %>%
    add_trace(m, x = ~Original, y = ~FDRcorrected, type = "scatter",
              mode="lines", line=list(color='grey')) %>%
    add_trace(m, x = ~logFDR.x, y = ~logFDR.y, type = "scatter", mode = "markers",
              showlegend = FALSE, text = m$Description,
              hoverinfo = 'text') %>%
    layout(title="log FDR comparison")
} else if (comparison == "Both errors") {
  m$Original <- m$BOTHcorrected <- round(VEC[1:nrow(m)],0.1)
  m$logFDR.x <- signif(m$logFDR.x,3)
  m$logFDR.y <- signif(m$logFDR.y,3)

  plot_ly(
    m, x = ~logFDR.x, y = ~logFDR.y
  ) %>%
    add_trace(m, x = ~Original, y = ~BOTHcorrected, type = "scatter",
              mode="lines", line=list(color='grey')) %>%
    add_trace(m, x = ~logFDR.x, y = ~logFDR.y, type = "scatter", mode = "markers",
              showlegend = FALSE, text = m$Description,
              hoverinfo = 'text') %>%
    layout(title="log FDR comparison")
}
